Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-08 paper/main.tex

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    \uIf{ $(V,E \cup L \cup e)$ is acyclic }{
            	$L \gets L \cup  \{ e \} $\;
	    } \Else {
            	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    }
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

@@

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected 
in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ like when using a strong ancestral oracle; 
since only correctly oriented
edges are added between ancestor-descendant pairs, 
such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, 
an arbitrary edge is added in a manner that preserves acyclicity.
The result is a supergraph of $G^+$.
\end{proof}

@@

\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Prune}($V$,$E$,$\mathcal{D}_G$)}{
    let $L \gets \{\}$\;
    \ForEach{$X \to Y \in E$}{
        remove $X \to Y$ from $E$\;
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 1$}{
            $L \gets L \cup  \{X \to Y\}$ \;
        }
    }
    remove all edges in $L$ from $G$\;
    \Return{$(V,E)$}\;
}
\caption{Pruning superfluous edges}
\label{algo:prune}
\end{algorithm}

>>>>

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    $L \gets L \cup \{ e \} $ \;
	    \If{ $ (V, E \cup L) $ is cyclic }{
		    $ L \gets \textsc{FixCycles}(V, E, L, e, \mathcal{D}_G) $ \;
	    }
	    % \If{ $(V,E \cup L \cup e)$ is acyclic }{
            % 	$L \gets L \cup  \{ e \} $\;
	    % } \Else {
            % 	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    % }
        }
	\If{$|L|\geq k$}{\KwGoTo 16\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{FixCycle}{FixCycle}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc FixCycles}($V, E, L, e, \mathcal{D}_G $)}{
	$ u, v \gets e $ \;
	\ForEach {$ P = \{ v \rightarrow x_1, x_1 \rightarrow x_2, \cdots,  x_{k-1} \rightarrow x_k, x_k \rightarrow u \} $ in ($V, E \cup L$) }{
		\ForEach {$ x_i \rightarrow x_j $ in $ P $}{
        	let $\mathbf{Z}$ be a set that $d$-separates $x_i$ and $x_j$ in $(V, E \cup L)$\;
		\If {$ \mathcal{D}_G(X, Y, \mathbf{Z}) = 0 $}{
			$ L \gets L - \{ x_i \rightarrow x_j \} $ \;

		}
	}
	}
	\Return $ L $ \;
}
\caption{Fixing cycles by removing incorrect edges.}
\label{algo:fix_cycles}
\end{algorithm}

@@

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected 
in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ like when using a strong ancestral oracle; 
since only correctly oriented
edges are added between ancestor-descendant pairs, 
such additions never create cycles.
For $X, Y$ that are only $d$-connected through common ancestors, 
an arbitrary edge is added in a manner that preserves acyclicity.
The result is a supergraph of $G^+$.
\end{proof}

@@

\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Prune}($V$,$E$,$\mathcal{D}_G$)}{
    let $L \gets \{\}$\;
    \ForEach{$X \to Y \in E$}{
        remove $X \to Y$ from $E$\;
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 1$}{
            $L \gets L \cup  \{X \to Y\}$ \;
        }
    }
    remove all edges in $L$ from $G$\;
    \Return{$(V, E)$}\;
}
\caption{Pruning superfluous edges}
\label{algo:prune}
\end{algorithm}
